Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-02-26 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/SM_deNovoRates.tex 	modified:   Manuscripts/deNovoRates.tex 	modified:   Manuscripts/refs.bib

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

@@

\begin{abstract}
New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence. To synthesize a protein, DNA must be transcribed as well as translated. Both processes need certain DNA sequence features. Stable transcription requires promoters and a polydenylation signal, while translation requires at least an open reading frame (ORF). We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost. We also investigate the effect of the order by which DNA features evolve, and if sequence composition is biased by mutation rate. We rationalize how genes are lost much more rapidly than they emerge, and how genes with long ORFs preferentially arise in regions that are already transcribed. Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
\end{abstract}

@@

$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/pGainLossFix2.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}. lack lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D. melanogaster} intergenic genome.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population \citep{KimuraFix}. To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals. First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$). The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$). We calculated the log-log ratio of the two probabilities under three scenarios (\autoref{pFix}), wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$). We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously. Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele. Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF. For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals. Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons. When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population.}

Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population.

Despite low probability of gene gain at one particular locus ($\sim 10^{-15}$ per generation), the overall rate of gene gain throughout the genome could be much higher. To this end we calculated the total gain rate of genes of any length, at any locus in \textit{D. melanogaster} intergenic genome. We found this rate to be $2.8\times 10^{-5}$ new genes per generation, that corresponds to approximately $7.5\times 10^{-5}$ new genes per year \citep[assuming \textit{D. melanogaster} generation time of two weeks;][]{drosophilaModelOrg}. We performed an analogous analysis on the rate of gain of new transcripts and found this rate to be 1.53 transcripts per year. This estimate is slightly higher than the rate of transcript gain predicted from \textit{D. melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}.

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whoisfirstX.pdf}
\caption{Proto-genes preferentially emerge RNA-first. The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(B)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}). A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}. In both panels, the horizontal axis denotes the number of codons in the ORF.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first). In other words, the RNA-first trajectory is more feasible. Likewise, a negative value of the ratio would suggest that the ORF-first trajectory is more feasible. We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2}). More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length. This is true for loci with three different values of GC content (34\%, 42\%, 50\%). However, our complementary analysis using DNA oligomer frequencies from \textit{D. melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length. Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.}

@@

\begin{figure}[!b]
\centering
\hrule
\vspace{1ex}
\includegraphics[scale=0.5]{Figures/pLoss_promoterlessX.pdf}
\caption{ORF loss in proto-genes is more probable than transcription loss. The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}. Black lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D. melanogaster} intergenic genome.}
\label{lossprob}
\end{figure}

@@

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge. In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C). Two evolutionary scenarios could explain this observation. In the first scenario, the proto-gene simply emerged for the first time in population A. The second scenario posits that the proto-gene was present in the ancestral population but was lost in both population B and population C. Our models suggest that the second scenario is more likely than is assumed. \cmnt{This inference can be further extended to the level of closely related species where the divergence times are relatively shorter than the timescale of multiple gain and loss events within a lineage}. Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss.

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first). Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

@@

We show that transcription loss is less likely than the loss of long ORFs. However, it may be difficult to find direct evidence for such a phenomenon. Specifically, it is difficult to infer from data if transcription was lost or never emerged. If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales. One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status. Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF. This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter. If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.

@@

Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. \cmnt{Since our model primarily focuses on gene emergence at one locus, it can be used to predict evolutionary probabilities for different genomic loci. Furthermore, several predictions of our models are not qualitatively affected by GC-content. More realistic nucleotide composition distribution estimated from \textit{D. melanogaster} intergenic sequences, also qualitatively agree with GC-content based predictions. Thus, most predictions are qualitatively robust to small differences in parameters. Second, in our models, we assumed uniform mutation rate throughout the genome. However, recent studies have shown Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}}. Our models can be adapted to investigate specific genomic loci, where estimates of the local nucleotide composition and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Third, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. Mutation bias may have more significant effect on the predictions than mutation rate. Existing data suggest that different species have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Fourth, we assume that occurrence of a point mutation is independent of another mutation. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Fourth, we did not specifically model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes. For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}. \cmnt{To model this phenomenon in our analyses, we assumed that DNA loci feasible for transcript emergence are close to known \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}. Specifically, we assumed that transcription can be initiated at a probability that is equal to the proportion of such regulatory elements in the intergenic genome. This is a rough estimate but it can be updated when more specific data becomes available. For example, ATAC-seq data can provide reliable estimates of nucleosome depleted regions, that can act as transcription initiation sites \citep{ATAC-seq}. This analysis can be supplemented with data from techniques like native elongating transcript sequencing \citep[NET-seq;][]{NET-seq} and global run-on sequencing \citep[GRO-seq;][]{GRO-seq}, that can detect nascent transcripts.} Our modeling framework also allows incorporation of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}). Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.

A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}. Our models are not an exception. They may not be 100\% accurate but they make useful predictions that can lead to a more focused experimental validation. \cmnt{Many of our predictions are qualitatively robust to small parameter changes. Some predictions are also in agreement with experimental data. For example, our prediction of transcript gain rate is only slightly inflated than the estimated rates using an empirical analysis of \textit{D. melanogaster} populations \citep{AnnaPeteTranscripts}. This inflated estimate could have possibly resulted from our relaxed requirements for a promoter, or non-inclusion of other unknown factors that may have affected population evolution. When supplemented with experimental data analysis our models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution, and can provide more accurate predictions.} Therefore our work opens up an opportunity for theoretical and computational biologists to design analyses using our modeling framework, that more accurately describe their system of interest.

@@

We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:

\begin{table}[H]
\centering
\begin{tabular}{c c c c}
\toprule
\textbf{Case} & \textbf{Gain of gene}  & \textbf{Loss of gene} & \textbf{Selection coefficient (s)}\\
\midrule
No selection on protogene & $\dfrac{1}{2N_e}$ & $\dfrac{1}{2N_e}$ & -- \\
\midrule
Selection on protogene & $\dfrac{1 - e^{-2s}}{1- e^{-4N_es}}$ & $\dfrac{1 - e^{2s}}{1- e^{4N_es}}$ & $\pm0.001$\\
\bottomrule
\end{tabular}
\caption{Gene fixation/extinction probabilities with or without selection. Here $N_e$ denotes effective population size and $s$ denotes the selection co-efficient denoting the fitness advantage provided by the protein encoded in the proto-gene (positive valued for a beneficial protein, and negative valued for a deleterious protein).}
\label{pFix}
\end{table}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec,lineno}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

@@

\linenumbers

\section*{Abstract}
New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence. To synthesize a protein, DNA must be transcribed as well as translated. Both processes need certain DNA sequence features. Stable transcription requires promoters and a polydenylation signal, while translation requires at least an open reading frame (ORF). We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost. We also investigate the effect of the order by which DNA features evolve, and if sequence composition is biased by mutation rate. We rationalize how genes are lost much more rapidly than they emerge, and how genes with long ORFs preferentially arise in regions that are already transcribed. Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.

@@

\begin{equation}
\text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})
\label{loglogratio}
\end{equation}

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/pGainLossFix2.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in all panels shows the number of independent gene loss events with extinction (\autoref{genelosseq}) that can occur relative to one gene gain event with fixation (\autoref{loglogratio}, \autoref{pFix}), under \textbf{(A)} no selection, \textbf{(B)} weak positive selection and \textbf{(C)} weak negative selection on the expression of the protein from a proto-gene. Horizontal axes show the number of codons in the ORF. Black lines in panel \textbf{(A)} denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D. melanogaster} intergenic genome. In panels \textbf{(B)} and \textbf{(C)} we use the probabilitiy values from \textit{D. melanogaster} intergenic genome. $N_e$ denotes effective population size and $s$ denotes the selection co-efficient.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population \citep{KimuraFix}. To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals. First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$, \autoref{pFix}). The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$, \autoref{pFix}). We calculated the log-log ratio of the two probabilities under three scenarios (\autoref{pFix}), wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$). We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously. Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF (\hyperref[gainlossprob]{\autoref{gainlossprob}B}). For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals. Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons. When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population (\hyperref[gainlossprob]{\autoref{gainlossprob}C}). We further note that time to fixation of a rare allele (for example, a newly emerged protogene) would be much smaller \citep[$4N_e$ generations;][]{KimuraFix} than the time to gain ($10^{14}$ -- $10^{20}$ generations) or lose it ($10^{7}$ -- $10^{8}$ generations).}

Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population. \cmnt{This is also true for a model of gene emergence that requires a core promoter (TATA-box or Inr) for initiating transcription ({\color{blue} Figure S1}).

Despite low probability of gene gain at one particular locus ($\sim 10^{-15}$ per generation), the overall rate of gene gain throughout the genome could be much higher. To this end we calculated the total gain rate of genes of any length, at any locus in \textit{D. melanogaster} intergenic genome. We found this rate to be $2.8\times 10^{-5}$ new genes per generation, that corresponds to approximately $7.5\times 10^{-5}$ new genes per year \citep[assuming \textit{D. melanogaster} generation time of two weeks;][]{drosophilaModelOrg}. We performed an analogous analysis on the rate of gain of new transcripts and found this rate to be 1.53 transcripts per year. This estimate is slightly higher than the rate of transcript gain predicted from \textit{D. melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}}.

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whoisfirstX.pdf}
\caption{Proto-genes preferentially emerge RNA-first. The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(B)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}). A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}. In both panels, the horizontal axis denotes the number of codons in the ORF. Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D. melanogaster} intergenic genome.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first). In other words, the RNA-first trajectory is more feasible. Likewise, a negative value of the ratio would suggest that the ORF-first trajectory is more feasible. We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length. This is true for loci with three different values of GC content (34\%, 42\%, 50\%). However, our complementary analysis using DNA oligomer frequencies from \textit{D. melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length. Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence. In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Inr) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2})}

@@

\begin{figure}[!b]
\centering
\hrule
\vspace{1ex}
\includegraphics[scale=0.5]{Figures/pLoss_promoterlessX.pdf}
\caption{ORF loss in proto-genes is more probable than transcription loss. The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}. Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D. melanogaster} intergenic genome.}
\label{lossprob}
\end{figure}

@@

We show that when a protein product of a proto-gene doesn't affect organismal fitness, the gene can be lost much more rapidly than they are gained, such that it can be independently lost two to three times in the time required for it to emerge (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C). Two evolutionary scenarios could explain this observation. In the first scenario, the proto-gene simply emerged for the first time in population A. The second scenario posits that the proto-gene was present in the ancestral population but was lost in both population B and population C. Our models suggest that the second scenario is more likely than is assumed. \cmnt{This inference can be further extended to the level of closely related species where the divergence times are relatively shorter than the timescale of multiple gain and loss events within a lineage}. Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss. \cmnt{Proto-genes that encode proteins that are beneficial to an organism's fitness are likely to be protected in the genome against gene loss \citep{Lee2019}. Our models indeed show that loss of even moderately beneficial genes is unlikely in large populations (\hyperref[gainlossprob]{\autoref{gainlossprob}B})}.

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first, \autoref{whoisfirst}). Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

@@

We show that transcription loss is less likely than the loss of long ORFs (\autoref{lossprob}). However, it may be difficult to find direct evidence for such a phenomenon. Specifically, it is difficult to infer from data if transcription was lost or never emerged. If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales. One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status. Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF. This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter. If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.

@@

Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. \cmnt{Since our model primarily focuses on gene emergence at one locus, it can be used to predict evolutionary probabilities for different genomic loci. Furthermore, several predictions of our models are not qualitatively affected by GC-content. More realistic nucleotide composition distribution estimated from \textit{D. melanogaster} intergenic sequences, also qualitatively agree with GC-content based predictions. Thus, most predictions are qualitatively robust to small differences in parameters. Second, in our models, we assumed uniform mutation rate throughout the genome. However, recent studies have shown that mutation rates are also not uniform across the genome, and can vary significantly \citep{mutbiasArabidopsis}}. Our models can be adapted to investigate specific genomic loci, where estimates of the local nucleotide composition and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Third, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. Mutation bias may have more significant effect on the predictions than mutation rate. Existing data suggest that different species have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Fourth, we assume that occurrence of a point mutation is independent of another mutation. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Fourth, we did not specifically model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes. For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}. \cmnt{To model this phenomenon in our analyses, we assumed that DNA loci feasible for transcript emergence are close to known \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}. Specifically, we assumed that transcription can be initiated at a probability that is equal to the proportion of such regulatory elements in the intergenic genome. This is a rough estimate but it can be updated when more specific data becomes available. For example, ATAC-seq data can provide reliable estimates of nucleosome depleted regions, that can act as transcription initiation sites \citep{ATAC-seq}. This analysis can be supplemented with data from techniques like native elongating transcript sequencing \citep[NET-seq;][]{NET-seq} and global run-on sequencing \citep[GRO-seq;][]{GRO-seq}, that can detect nascent transcripts.} Our modeling framework also allows incorporation of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}). Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.

A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}. Our models are not an exception. They may not be 100\% accurate but they make useful predictions that can lead to a more focused experimental validation. \cmnt{Many of our predictions are qualitatively robust to small parameter changes. Some predictions are also in agreement with experimental data. For example, our prediction of transcript gain rate is only slightly inflated than the estimated rates using an empirical analysis of \textit{D. melanogaster} populations \citep{AnnaPeteTranscripts}. This difference could have possibly resulted from our relaxed requirements for a promoter, or other unknown factors that may have affected population evolution \citep[for example, the actual effective population size and environmental conditions;][]{dmel3decades}. When supplemented with experimental data analysis our models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution, and can provide more accurate predictions.} Therefore our work opens up an opportunity for theoretical and computational biologists to design analyses using our modeling framework, that more accurately describe their system of interest.

@@

\newcommand{\specialcell}[2][c]{%
  \bfseries\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\cmnt{
We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:}

\begin{table}[H]
\centering
\begin{tabular}{c c c c}
\toprule
\textbf{Case} & \specialcell{Gain of gene\\(Fixation, $P_\textit{fix}$)}  & \specialcell{Loss of gene\\ (Extinction, $P_\textit{ext}$)} & \specialcell{Selection\\coefficient ($s$)}\\
\midrule
No selection on protogene & $\dfrac{1}{2N_e}$ & $\dfrac{1}{2N_e}$ & -- \\
\midrule
Selection on protogene & $\dfrac{1 - e^{-2s}}{1- e^{-4N_es}}$ & $\dfrac{1 - e^{2s}}{1- e^{4N_es}}$ & $\pm0.001$\\
\bottomrule
\end{tabular}
\caption{Gene fixation/extinction probabilities with or without selection. Here $N_e$ denotes effective population size and $s$ denotes the selection co-efficient denoting the fitness advantage provided by the protein encoded in the proto-gene (positive valued for a beneficial protein, and negative valued for a deleterious protein).}
\label{pFix}
\end{table}
